\bf Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-06-25 paper.qmd
index definition

To compare a new sample with an existing norm, like a multivariate normal distribution, in higher than two dimensions, we have typically used two samples of points. The norm is represented by points on the surface of a $p$-dimensional ellipse, corresponding to a confidence level. A sample of points uniformly distributed on a $p$-dimensional sphere is generated by

1. Simulating a sample of observations ($\mathbfit{x}$)
from $N_p(\mu, \Sigma)$. 
2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$.

To convert this to points on the surface of a confidence ellipse,

@@

Finally, new observations can be visually compared with this ellipse by

@@

```{r}
#| echo: false
#| label: fig-ci
#| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1. A confidence ellipse is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipse (d)."
#| out-width: 80%
set.seed(717)
d1 <- matrix(rnorm(716 * 2), ncol = 2) 
d2 <- t(apply(d1, 1, geozoo:::l2norm_vec))
vc <- matrix(c(1, -0.5, -0.5, 1), ncol=2, byrow=TRUE)
evc <- eigen(vc)
vc2 <- (evc$vectors) %*% diag(sqrt(evc$values)) %*% t(evc$vectors)
d3 <- d2 %*% vc2 |>
  as.data.frame()

@@

p1 <- ggplot(d1, aes(x=V1, y=V2)) +
  geom_point() +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  ggtitle("a. Bivariate standard normal sample") +
  theme(axis.text = element_blank())

p2 <- ggplot(d2, aes(x=V1, y=V2)) +
  geom_point() +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  ggtitle("b. Uniform on a sphere") +
  theme(axis.text = element_blank())

p3 <- ggplot(d3, aes(x=V1, y=V2)) +
  geom_point() +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  ggtitle("c. Points on surface of confidence ellipse.") +
  theme(axis.text = element_blank())

p4 <- ggplot(d4, aes(x=V1, y=V2, shape=type)) +
  geom_point() +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  ggtitle("d. Observations inside and outside confidence ellipse.") +
  theme(legend.position = "none", 
        axis.text = element_blank())

p1 + p2 + p3 + p4 + plot_layout(ncol=2)
```

@@

>>>>

To compare a new sample with an existing norm, like a multivariate normal distribution, in higher than two dimensions, we have typically used two samples of points. The norm is represented by points on the surface of a $p$-dimensional ellipsoid, corresponding to a confidence level. A sample of points uniformly distributed on a $p$-dimensional sphere is generated by

1. Simulating a sample of observations ($\mathbfit{x}$, which are \pD{} vectors)
from $N_p(\mathbfit{\mu}, \Sigma)$. 
2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$.

To convert this to points on the surface of a confidence ellipsoid,

@@

Finally, new observations can be visually compared with this ellipsoid by

@@

```{r}
#| echo: false
#| label: fig-ci
#| fig-width: 8
#| fig-height: 2
#| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1. A confidence ellipsoid is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipsoid (d)."
set.seed(717)
d1 <- matrix(rnorm(716 * 2), ncol = 2) 
d2 <- t(apply(d1, 1, geozoo:::l2norm_vec))
vc <- matrix(c(1, -0.5, -0.5, 1), ncol=2, byrow=TRUE)
evc <- eigen(vc)
vc2 <- (evc$vectors) %*% diag(sqrt(evc$values)) %*% t(evc$vectors)
d3 <- d2 %*% vc2 |>
  as.data.frame()

@@

p1 <- ggplot(d1, aes(x=V1, y=V2)) +
  geom_point(alpha=0.5) +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  xlim(c(-3.2, 3.2)) +
  ylim(c(-3.2, 3.2)) +
  interior_annotation("a") +
  #ggtitle("a. Bivariate standard normal sample") +
  theme(axis.text = element_blank())

p2 <- ggplot(d2, aes(x=V1, y=V2)) +
  geom_point(alpha=0.5) +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  xlim(c(-1.2, 1.2)) +
  ylim(c(-1.2, 1.2)) +
  interior_annotation("b") +
  #ggtitle("b. Uniform on a sphere") +
  theme(axis.text = element_blank())

p3 <- ggplot(d3, aes(x=V1, y=V2)) +
  geom_point(alpha=0.5) +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  xlim(c(-1.2, 1.2)) +
  ylim(c(-1.2, 1.2)) +
  interior_annotation("c") +
  #ggtitle("c. Points on surface of confidence ellipsoid.") +
  theme(axis.text = element_blank())

p4 <- ggplot(d4, aes(x=V1, y=V2, shape=type)) +
  geom_point(alpha=0.5) +
  xlab(expression(x[1])) +
  ylab(expression(x[2])) +
  xlim(c(-1.2, 1.2)) +
  ylim(c(-1.2, 1.2)) +
  interior_annotation("d") +
  #ggtitle("d. Observations inside and outside confidence ellipsoid.") +
  theme(legend.position = "none", 
        axis.text = element_blank())

p1 + p2 + p3 + p4 + plot_layout(ncol=4)
```

<PLOTS FROM USCHI's PHYSICS EXAMPLE HERE>

Although this is flexible, this does not make it easy to guide the tour towards the directions (projections) where the samples are most different from the normal. What would be desirable is to analytically define the confidence ellipsoid, compute flag observations that are outside, steer the tour to projections that reveal the extent of the difference. And also display the projected ellipsoid as a geometric shape rather than a sample of points. These are the procedures that are described in the next section.

@@

## Projecting an ellipsoid

A \pD{} ellipsoid corresponding to a given variance-covariance ($\Sigma$) is defined by

$$
(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = c^2
$$
where $c$ a constant that depends on a specific confidence level.

The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane. We call points in the projection that are on the curve $\mathbfit{y}$. From this we can compute the analog equation for the projection as

$$(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2$$
where $P$ is a $(p\times 2)$ orthonormal basis defining the projection and $\mathbfit{\mu}_p = \mathbfit{\mu} P$ the projected mean. This means the matrix $(P^T \Sigma P)^{-1}$ is defining the ellipse in the \gD{} projection. In general $c$ could be any constant, but typically we would select it as a quantile of the $\chi^2$ distribution, so that the size of the ellipse corresponds to a selected probability.

<IMAGES OF PROJECTED ELLIPSE>

### Index specification

To define a measure of an interesting projection is to maximize the **average Malahanobis distance in the projection** for a subset of points, $W$. The set of points could be chosen in different ways, but the default is those that are outside the specified ellipsoid in \pD{}. Alternatives could be to select a set of observations with the largest Mahalanobis distance, manually select observations or possibly a group of points identified via clustering of the extremes.

The index is written as

$$
\sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T
$$

where by default $W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}$, is the set of observations outside the \pD{} ellipsoid.

## Implementation

This was implemented in the `tourr` (REF) package, where the projected ellipsoid can be drawn for each projection.

### 2024-06-25 paper.tex
index definition

To compare a new sample with an existing norm, like a multivariate
normal distribution, in higher than two dimensions, we have typically
used two samples of points. The norm is represented by points on the
surface of a \(p\)-dimensional ellipse, corresponding to a confidence
level. A sample of points uniformly distributed on a \(p\)-dimensional
sphere is generated by

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulating a sample of observations (\(\mathbfit{x}\)) from
  \(N_p(\mu, \Sigma)\).
\item
  Transforming each observation to have unit distance from the mean,
  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\).
\end{enumerate}

To convert this to points on the surface of a confidence ellipse,

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-ci-1.pdf}

@@

\caption{\label{fig-ci}Simulating a uniform sample on a sphere involves
sampling from a multivariate normal (a) and transforming each
observation to have length equal to 1. A confidence ellipse is generated
by transforming the sphere relative to a specified variance-covariance
matrix (c), and new observations can be visually assessed to be inside
or outside by plotting with the ellipse (d).}

@@

@@

>>>>

To compare a new sample with an existing norm, like a multivariate
normal distribution, in higher than two dimensions, we have typically
used two samples of points. The norm is represented by points on the
surface of a \(p\)-dimensional ellipsoid, corresponding to a confidence
level. A sample of points uniformly distributed on a \(p\)-dimensional
sphere is generated by

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulating a sample of observations (\(\mathbfit{x}\), which are \pD{}
  vectors) from \(N_p(\mathbfit{\mu}, \Sigma)\).
\item
  Transforming each observation to have unit distance from the mean,
  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\).
\end{enumerate}

To convert this to points on the surface of a confidence ellipsoid,

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-ci-1.pdf}

@@

\caption{\label{fig-ci}Simulating a uniform sample on a sphere involves
sampling from a multivariate normal (a) and transforming each
observation to have length equal to 1. A confidence ellipsoid is
generated by transforming the sphere relative to a specified
variance-covariance matrix (c), and new observations can be visually
assessed to be inside or outside by plotting with the ellipsoid (d).}

@@

\textless PLOTS FROM USCHI's PHYSICS EXAMPLE HERE\textgreater{}

Although this is flexible, this does not make it easy to guide the tour
towards the directions (projections) where the samples are most
different from the normal. What would be desirable is to analytically
define the confidence ellipsoid, compute flag observations that are
outside, steer the tour to projections that reveal the extent of the
difference. And also display the projected ellipsoid as a geometric
shape rather than a sample of points. These are the procedures that are
described in the next section.

@@

\section{Projecting an ellipsoid}\label{projecting-an-ellipsoid}

A \pD{} ellipsoid corresponding to a given variance-covariance
(\(\Sigma\)) is defined by

\[
(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = c^2
\] where \(c\) a constant that depends on a specific confidence level.

The projection of an ellipsoid onto \gD{} is an ellipse, where the curve
of the ellipse is defined through the set of points \(\mathbfit{x}\) for
which the gradient is parallel to the projection plane. We call points
in the projection that are on the curve \(\mathbfit{y}\). From this we
can compute the analog equation for the projection as

\[(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2\]
where \(P\) is a \((p\times 2)\) orthonormal basis defining the
projection and \(\mathbfit{\mu}_p = \mathbfit{\mu} P\) the projected
mean. This means the matrix \((P^T \Sigma P)^{-1}\) is defining the
ellipse in the \gD{} projection. In general \(c\) could be any constant,
but typically we would select it as a quantile of the \(\chi^2\)
distribution, so that the size of the ellipse corresponds to a selected
probability.

\subsection{Index specification}\label{index-specification}

To define a measure of an interesting projection is to maximize the
\textbf{average Malahanobis distance in the projection} for a subset of
points, \(W\). The set of points could be chosen in different ways, but
the default is those that are outside the specified ellipsoid in \pD{}.
Alternatives could be to select a set of observations with the largest
Mahalanobis distance, manually select observations or possibly a group
of points identified via clustering of the extremes.

The index is written as

\[
\sum_{\mathbfit{w} \in W} (\mathbfit{w} - \mathbfit{\mu}) P (P^T\Sigma P)^{-1}P^T(\mathbfit{w} - \mathbfit{\mu})^T
\]

where by default
\(W = \{\mathbfit{x}: (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T > c^2\}\),
is the set of observations outside the \pD{} ellipsoid.

\section{Implementation}\label{implementation}

This was implemented in the \texttt{tourr} (REF) package, where the
projected ellipsoid can be drawn for each projection.
